---
title: "Inter-compartmental interactions (A/B ratio)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(GenomicRanges)
library(foreach)
library(doParallel)
library(dplyr)
library(ggplot2)
library(stringr)
library(bsseq)
library(Matrix)
library(reshape2)
library(tidyverse)
library(cowplot)

date()
bin_size <- "100000"
chrs <- paste0("chr", 1:22)

scaleCols <- c("#4daf4a", "orange", "#984ea3")
names( scaleCols ) <- c("Normal", "Adenoma", "Tumor")
scaleCol <- scale_color_manual(values=scaleCols)
scaleFill <- scale_fill_manual(values=scaleCols)
comp_a_col <- "#0002aa"
comp_b_col <- "#f7cd46"
comp_i_col <- "#66BFE3"
scaleCompCols <- c(comp_a_col, comp_i_col, comp_b_col)
names( scaleCompCols ) <- c("A", "I", "B")
scaleCompCol <- scale_color_manual(values=scaleCompCols)
scaleCompFill <- scale_fill_manual(values=scaleCompCols)
```


# Tumor/Normal: Prepare a long format A/B ratio data frame
```{r }
comp <- readRDS("../rds/subcompartment_labels_annotate.rds")
meth_gr <- readRDS("../rds/compartmentGR.rds")

# Not used: 'MGH1904-sb','MGH1904-sb2','BRD3170N-sb','BRD3462N-sb','BRD3462N-sb2','BRD3162N-sb','BRD3409N-sb','BRD3409N-sb2'
hic_samples <- c('BRD3179N','BRD3187N','BRD3328N','BRD3462N','BRD3170N-sb2','BRD3162N-sb2','BRD3409N',
                 'Adenoma0410T',
                 'BRD3179','BRD3187','MGH1904','MGH8416','BRD3162','MGH2834','MGH5328','MGH2231-sb2','MGH3535','BRD3462','BRD3412','BRD3378') 

validation <- c("BRD3170N", "BRD3162N", "BRD3409N",
                "BRD3378", "BRD3412", "BRD3462", "MGH3535")

x <- colnames(mcols(meth_gr))
meth <- mcols(meth_gr)[,x[grepl("rep1_opensea", x)]]
colnames(meth) <- sub("_rep1_opensea", "", colnames(meth))
#colMeans(!is.na(meth))

rds <- paste0("../rds/ab_interactions_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {
  registerDoParallel(cores=4)
  df <- foreach(sample = hic_samples, .combine=rbind) %dopar% {
    sparse_rds <- paste0("../terra/", sample, ".sparsehic.rds")
    # Drop the -sb* suffix for the low coverage samples for matching with methylation
    sample <- sub("-sb.*", "", sample)
    x <- readRDS( sparse_rds )@resolutionNamedList[[bin_size]]
    df2 <- foreach (chr=chrs, .combine=rbind) %do% {
      #chr <- "chr1"
      mat <- as.matrix(x[[chr]])
      mat_oe <- melt(mat, varnames = c("x", "y"), value.name = "obs") %>% filter(y>=x) %>% 
        mutate(d=y-x) %>%
        group_by(d) %>% mutate(obs_exp = obs / mean(obs, na.rm=T)) %>% ungroup() %>% 
        select(x,y,obs_exp) %>% spread(y,obs_exp) %>% column_to_rownames("x") %>%
        as.matrix()
      
      mat[lower.tri(mat)] = t(mat)[lower.tri(mat)]
      mat_oe[lower.tri(mat_oe)] = t(mat_oe)[lower.tri(mat_oe)]
      
      gr <- unlist(tile(GRanges(chr, IRanges(1, ncol(mat) * as.numeric(bin_size))), width=as.numeric(bin_size)))
      idx <- findOverlaps(gr, comp, select = "first")
      gr$compartment <- comp$subComp[idx]

      # >1Mb interactions
      mat <- as.matrix(mat - band(mat, -10, 10))
      gr$a_ix_sum_1mb <- colSums(mat[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_sum_1mb <- colSums(mat[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_sum_1mb <- colSums(mat[gr$compartment=="B",], na.rm=TRUE)
      
      # O/E
      gr$a_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="B",], na.rm=TRUE)
      
      idx <- findOverlaps(gr, meth_gr, select = "first")
      if (sample %in% colnames(meth)) {
        gr$meth <- meth[idx, sample]  
      } else {
        gr$meth <- NA
      }
      
      df <- gr %>% as.data.frame() %>% select(seqnames, start, compartment, meth,
                                        a_ix_sum_1mb, i_ix_sum_1mb, b_ix_sum_1mb,
                                        a_ix_mean_oe, i_ix_mean_oe, b_ix_mean_oe) %>% 
        cbind(sample=sample, type=ifelse(grepl("N", sample), "Normal", "Tumor"))
      df
    }
  }  
  df$type <- as.character(df$type)
  df$type=ifelse(grepl("N", df$sample), "Normal", "Tumor")
  df$type[grepl("Adenoma", df$sample)] <- "Adenoma"
  df$type <- relevel(factor(df$type), ref="Normal")
  df$compartment <- factor(df$compartment, levels=c("A", "I", "B"))
  df$cohort = ifelse(df$sample %in% validation, "Cohort 2", "Cohort 1")
  saveRDS(df, file=rds)
}

# Drop single adenoma sample
df <- df %>% filter(type != "Adenoma")

```



# Tumor vs Normal A/B ratio (O/E, Both cohorts together)

Each data point represents 1 sample
```{r}
# Compartments A,B O/E
set.seed(124)
df %>% filter(compartment %in% c("A", "B")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_ab.png", width=4, height=3)

# Compartments I O/E
set.seed(124)
df %>% filter(compartment %in% c("I")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_i.png", width=2.5, height=3)

```


# Tumor vs Normal A/B ratio by cohort (O/E)

Each data point represents 1 sample
```{r}
# Compartments A,B O/E - Cohort 1
set.seed(124)
df %>% filter(cohort=="Cohort 1" & compartment %in% c("A", "B")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_ab_cohort1.png", width=3, height=3)

# Compartments A,B O/E - Cohort 2
df %>% filter(cohort=="Cohort 2" & compartment %in% c("A", "B")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_ab_cohort2.png", width=3, height=3)


# Compartment I O/E - Cohort 1
set.seed(124)
df %>% filter(cohort=="Cohort 1" & compartment %in% c("I")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_i_cohort1.png", width=2.5, height=3)

# Compartment I O/E - Cohort 2
set.seed(124)
df %>% filter(cohort=="Cohort 2" & compartment %in% c("I")) %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_oe=mean(ab_ratio_oe, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_oe, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio (O/E)") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_normal_tumor_one_point_per_sample_comp_i_cohort2.png", width=2.5, height=3)

```

# Tumor vs Normal A/B ratio (Observed matrix)

The versions below are the same as above but use the observed interaction matrix instead of O/E. (They look very similar)

```{r}
# Compartments A,B
set.seed(124)
df %>% filter(compartment %in% c("A", "B")) %>% mutate(ab_ratio_1mb = log2(a_ix_sum_1mb/b_ix_sum_1mb)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_1mb=mean(ab_ratio_1mb, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_1mb, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/extra/abratio_normal_tumor_one_point_per_sample_comp_ab.png", width=4, height=3)

# Compartment I
set.seed(124)
df %>% filter(compartment %in% c("I")) %>% mutate(ab_ratio_1mb = log2(a_ix_sum_1mb/b_ix_sum_1mb)) %>%
  group_by(type, sample, compartment) %>%
  summarize(ab_ratio_1mb=mean(ab_ratio_1mb, na.rm=TRUE)) %>%
  ggplot(aes(type, ab_ratio_1mb, fill=type)) + geom_jitter(aes(color=type), size=2) + geom_boxplot(alpha=0.1, aes(color=type), outlier.colour = NA)  + facet_wrap(~compartment, nrow=1, scales="free_y") + theme_bw() +scaleFill + ylab("Log2 A/B Interaction Ratio") + theme(legend.position = "none") + 
  scaleCol + xlab("") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/extra/abratio_normal_tumor_one_point_per_sample_comp_i.png", width=2.5, height=3)

```


# A/B ratio difference (O/E) vs hypomethylation relative to minimally hypomethylated regions

ab_ratio_oe = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe), chr_pos = paste(seqnames, start)) %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[type=="Normal"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25, 0.3,  1), labels=c("<15%", "15%-20%", "20%-25%", "25%-30%", ">30%")))

df_b <- df2 %>% filter(type=="Tumor" & seqnames=="chr1" & compartment=="B") 
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(type=="Tumor" & seqnames=="chr1" & compartment=="I") 
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name)) %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("A/B Ratio Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_by_hypomethylation_relative_to_minimally_hypometh_tumor.png", width=7, height=4)
```



# HCT116 AZA/DMSO: Prepare a long format A/B ratio data frame
```{r }
comp <- readRDS("../rds/granges_compI_HCT116.rds")
meth_gr <- readRDS("../rds/compartmentGR.rds")

samples <- c('HCT116-5AZA-24h-1','HCT116-5AZA-24h-2','HCT116-DMSO-24h-1','HCT116-DMSO-24h-2')
x <- colnames(mcols(meth_gr))
meth <- mcols(meth_gr)[,x[grepl("mean_opensea", x)]]
colnames(meth) <- sub("_mean_opensea", "", colnames(meth))

# Apply the same methylation values to the 2 HiC reps
colnames(meth)[colnames(meth)=="HCT116.5AZA.24h"] <- "HCT116-5AZA-24h-1"
colnames(meth)[colnames(meth)=="HCT116.DMSO.24h"] <- "HCT116-DMSO-24h-1"
meth["HCT116-5AZA-24h-2"] <- meth["HCT116-5AZA-24h-1"]
meth["HCT116-DMSO-24h-2"] <- meth["HCT116-DMSO-24h-1"]

rds <- paste0("../rds/ab_interactions_hct116_aza_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {
  registerDoParallel(cores=4)
  df <- foreach(sample = samples, .combine=rbind) %dopar% {
    sparse_rds <- paste0("../terra/", sample, ".sparsehic.rds")
    x <- readRDS( sparse_rds )@resolutionNamedList[[bin_size]]
    df2 <- foreach (chr=chrs, .combine=rbind) %do% {
      #chr <- "chr1"
      
      mat <- as.matrix(x[[chr]])
      mat_oe <- melt(mat, varnames = c("x", "y"), value.name = "obs") %>% filter(y>=x) %>% 
        mutate(d=y-x) %>%
        group_by(d) %>% mutate(obs_exp = obs / mean(obs, na.rm=T)) %>% ungroup() %>% 
        select(x,y,obs_exp) %>% spread(y,obs_exp) %>% column_to_rownames("x") %>%
        as.matrix()
      
      mat[lower.tri(mat)] = t(mat)[lower.tri(mat)]
      mat_oe[lower.tri(mat_oe)] = t(mat_oe)[lower.tri(mat_oe)]
      
      gr <- unlist(tile(GRanges(chr, IRanges(1, ncol(mat) * as.numeric(bin_size))), width=as.numeric(bin_size)))
      idx <- findOverlaps(gr, comp, select = "first")
      
      gr$compartment <- comp$subComp_HCT116[idx]

      # >1Mb
      mat <- as.matrix(mat - band(mat, -10, 10))
      gr$a_ix_sum_1mb <- colSums(mat[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_sum_1mb <- colSums(mat[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_sum_1mb <- colSums(mat[gr$compartment=="B",], na.rm=TRUE)
      
      #  O/E
      gr$a_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="B",], na.rm=TRUE)

      idx <- findOverlaps(gr, meth_gr, select = "first")
      if (sample %in% colnames(meth)) {
        gr$meth <- meth[idx, sample]  
      } else {
        gr$meth <- NA
      }
      
      rep <- str_split_fixed(sample, "-", 4)[4]
      
      gr %>% as.data.frame() %>% select(seqnames, start, compartment, meth,
                                        a_ix_sum_1mb, i_ix_sum_1mb, b_ix_sum_1mb,
                                        a_ix_mean_oe, a=i_ix_mean_oe, b_ix_mean_oe) %>% 
        cbind(sample=sample, rep=rep, type=ifelse(grepl("AZA", sample), "AZA", "DMSO"))
    }
  }  
  df$type <- relevel(df$type, ref="DMSO")
  df$compartment <- factor(df$compartment, levels=c("A", "I", "B"))
  saveRDS(df, file=rds)
}
```


# HCT116 Aza A/B ratio difference (O/E) relative to minimally hypomethylated regions

ab_ratio_oe = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe), chr_pos = paste(seqnames, start)) %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[type=="DMSO"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25,  1), labels=c("<15%", "15%-20%", "20%-25%", ">25%")))

df_b <- df2 %>% filter(compartment=="B") 
# Note: Convergence warning but max|grad| is low at 0.00334054
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name))
coeff$name <- factor(coeff$name, levels=unique(coeff$name))
 
coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("A/B Ratio Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_by_hypomethylation_relative_to_minimally_hypometh_hct116.png", width=7, height=4)
```

 



# Aging WI-38 fibroblasts: Prepare a long format A/B ratio data frame
```{r }
samples <- c('Wi38_16_1','Wi38_16_2','Wi38_30_1','Wi38_30_2','Wi38_40_1','Wi38_40_2','Wi38_46_1','Wi38_46_2')

rds <- paste0("../rds/ab_interactions_wi38_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {
  # Define compartments based on P16 eigenvector and P40-P16 hypomethylation
  comp <- readRDS("../rds/compartmentGR.rds")
  wgbs <- readRDS("../rds/wgbs_iw48.rds")
  tile_meth <- getMeth(wgbs, regions = comp, type = "raw", what = "perRegion")
  # Quantile normalize  wi_38_p40_1 to  wi_38_p40_2 because of skewed distribution
  o <- order(tile_meth[,"wi_38_p40_1"])
  tile_meth[,"wi_38_p40_1"] <- sort(tile_meth[,"wi_38_p40_2"])[o]
  comp$wi38_hypometh <- rowMeans(tile_meth[,c("wi_38_p16_1", "wi_38_p16_2")]) - tile_meth[,c("wi_38_p40_2")]
  #comp$wi38_hypometh <- rowMeans(tile_meth[,c("wi_38_p16_1", "wi_38_p16_2")]) - rowMeans(tile_meth[,c("wi_38_p40_1", "wi_38_p40_2")])
  comp$wi38_comp <- ifelse(comp$Wi38_16_mean_eigen>0, "A", "B")
  comp$wi38_comp[comp$Wi38_16_mean_eigen>0 & comp$wi38_hypometh>0.1] <- "I"
  #comp$wi38_comp[comp$Wi38_16_mean_eigen>0 & comp$wi38_hypometh_p30>0.1] <- "I"
  table(comp$wi38_comp)
  
  time <- sub("p", "", str_split_fixed(sampleNames(wgbs), "_", 4)[,3])
  rep <- str_split_fixed(sampleNames(wgbs), "_", 4)[,4]
  sampleNames(wgbs) <- paste0("Wi38_", time, "_", rep)
  
  registerDoParallel(cores=4)
  df <- foreach(sample = samples, .combine=rbind) %dopar% {
    #sample = samples[1] 
    sparse_rds <- paste0("../terra/", sample, ".sparsehic.rds")
    x <- readRDS( sparse_rds )@resolutionNamedList[[bin_size]]
    foreach (chr=chrs, .combine=rbind) %do% {
      mat <- as.matrix(x[[chr]])
      mat_oe <- melt(mat, varnames = c("x", "y"), value.name = "obs") %>% filter(y>=x) %>% 
        mutate(d=y-x) %>%
        group_by(d) %>% mutate(obs_exp = obs / mean(obs, na.rm=T)) %>% ungroup() %>% 
        select(x,y,obs_exp) %>% spread(y,obs_exp) %>% column_to_rownames("x") %>%
        as.matrix()
      
      mat[lower.tri(mat)] = t(mat)[lower.tri(mat)]
      mat_oe[lower.tri(mat_oe)] = t(mat_oe)[lower.tri(mat_oe)]
      
      gr <- unlist(tile(GRanges(chr, IRanges(1, ncol(mat) * as.numeric(bin_size))), width=as.numeric(bin_size)))
      idx <- findOverlaps(gr, comp, select = "first")
      gr$compartment <- comp$wi38_comp[idx]

      # >1Mb
      mat <- as.matrix(mat - band(mat, -10, 10))
      gr$a_ix_sum_1mb <- colSums(mat[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_sum_1mb <- colSums(mat[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_sum_1mb <- colSums(mat[gr$compartment=="B",], na.rm=TRUE)

      # O/E
      gr$a_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="A",], na.rm=TRUE)
      gr$i_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="I",], na.rm=TRUE)
      gr$b_ix_mean_oe <- colMeans(mat_oe[gr$compartment=="B",], na.rm=TRUE)
      
      tile_meth <- getMeth(wgbs, regions = gr, type = "raw", what = "perRegion")
      
      gr %>% as.data.frame() %>% select(seqnames, start, compartment, 
                                        a_ix_sum_1mb, i_ix_sum_1mb, b_ix_sum_1mb,
                                        a_ix_mean_oe, i_ix_mean_oe, b_ix_mean_oe) %>% 
        cbind(meth=tile_meth[,sample], sample=sample, time=str_split_fixed(sample, "_", 3)[,2], rep=str_split_fixed(sample, "_", 3)[,3])
    }
  }  
  
  saveRDS(df, file=rds)
}
```


# WI-38 fibroblasts P40 vs P16: A/B ratio difference (O/E) relative to minimally hypomethylated regions

ab_ratio_oe = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe), chr_pos = paste(seqnames, start)) %>%
  filter(time %in% c("16", "40") & sample!="Wi38_40_1") %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[time=="16"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25, 1), labels=c("<15%", "15%-20%", "20%-25%", ">25%"))) %>%
  ungroup()

df_b <- df2 %>% filter(compartment=="B") 
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(ab_ratio_oe ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name))
coeff$name <- factor(coeff$name, levels=unique(coeff$name))
 
coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("A/B Ratio Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_by_hypomethylation_relative_to_minimally_hypometh_wi38_p40_vs_p16.png", width=7, height=4)
```



# Aging WI-38 fibroblasts by time: A/B ratio relative to early passage

```{r}
df2 <- df %>% mutate(ab_ratio_oe = log2(a_ix_mean_oe/b_ix_mean_oe), chr_pos = paste(seqnames, start)) %>%
  filter(time %in% c("16", "30", "40") & sample!="Wi38_40_1") %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation40 = mean(meth[time=="16"], na.rm=TRUE) - mean(meth[time=="40"], na.rm=TRUE)) %>% ungroup() %>%
  filter(hypomethylation40 > 0.2)

df_b <- df2 %>% filter(compartment=="B") 
fit <- lmer(ab_ratio_oe ~ time + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(ab_ratio_oe ~  time + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("time", name))%>% mutate(name = sub("time", "P", name))

coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("WI-38 Passage Number") + ylab("A/B Ratio Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/abratio_oe_by_time_relative_to_p16_wi38_p40_hypometh_regions.png", width=7, height=4)

```




