---
title: "Inter-compartmental interactions (Eigenvector method)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(GenomicRanges)
library(foreach)
library(doParallel)
library(dplyr)
library(ggplot2)
library(stringr)
library(bsseq)
library(Matrix)
library(reshape2)
library(tidyverse)
library(cowplot)

date()
bin_size <- "100000"
chrs <- paste0("chr", 1:22)

scaleCols <- c("#4daf4a", "#984ea3")
names( scaleCols ) <- c("Normal", "Tumor")
scaleCol <- scale_color_manual(values=scaleCols)
scaleFill <- scale_fill_manual(values=scaleCols)
comp_a_col <- "#0002aa"
comp_b_col <- "#f7cd46"
comp_i_col <- "#66BFE3"
scaleCompCols <- c(comp_a_col, comp_i_col, comp_b_col)
names( scaleCompCols ) <- c("A", "I", "B")
scaleCompCol <- scale_color_manual(values=scaleCompCols)
scaleCompFill <- scale_fill_manual(values=scaleCompCols)
```


# Tumor/Normal: Prepare a long format eigen/meth data frame
```{r }
comp <- readRDS("../rds/subcompartment_labels_annotate.rds")
meth_gr <- readRDS("../rds/compartmentGR.rds")

# Not used: 'MGH1904-sb','MGH1904-sb2','BRD3170N-sb','BRD3462N-sb','BRD3462N-sb2','BRD3162N-sb','BRD3409N-sb','BRD3409N-sb2'
hic_samples <- c('BRD3179N','BRD3187N','BRD3328N','BRD3462N','BRD3170N-sb2','BRD3162N-sb2','BRD3409N',
                 'BRD3179','BRD3187','MGH1904','MGH8416','BRD3162','MGH2834','MGH5328','MGH2231-sb2','MGH3535','BRD3462','BRD3412','BRD3378') 

x <- colnames(mcols(meth_gr))
meth <- mcols(meth_gr)[,x[grepl("rep1_opensea", x)]]
colnames(meth) <- sub("_rep1_opensea", "", colnames(meth))
eigen <- mcols(meth_gr)[,x[grepl("rep1_eigen", x)]]
colnames(eigen) <- sub("_rep1_eigen", "", colnames(eigen))
colMeans(!is.na(meth))

rds <- paste0("../rds/eigen_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {
  registerDoParallel(cores=4)
  df <- foreach(sample = hic_samples, .combine=rbind) %dopar% {
    # Drop the -sb* suffix for the low coverage samples for matching with methylation
    sample <- sub("-sb.*", "", sample)
    df2 <- foreach (chr=chrs, .combine=rbind) %do% {
      idx <- seqnames(granges(meth_gr))==chr
      gr <- granges(meth_gr)[idx]
      
      if (sample %in% colnames(eigen)) {
        gr$eigen <- eigen[idx, sample]  
      } else {gr$eigen <- NA}
      if (sample %in% colnames(meth)) {
        gr$meth <- meth[idx, sample]  
      } else {gr$meth <- NA}
      
      idx <- findOverlaps(gr, comp, select = "first")
      gr$compartment <- comp$subComp[idx]

      gr %>% as.data.frame() %>% select(seqnames, start, compartment, eigen, meth) %>% 
        cbind(sample=sample, type=ifelse(grepl("N", sample), "Normal", "Tumor"))
    }
  }  
  df$type <- relevel(df$type, ref="Normal")
  df$compartment <- factor(df$compartment, levels=c("A", "I", "B"))
  saveRDS(df, file=rds)
}

```


# PC1 difference vs hypomethylation relative to minimally hypomethylated regions

eigen = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(chr_pos = paste(seqnames, start)) %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[type=="Normal"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25, 1), labels=c("<15%", "15%-20%", "20%-25%", ">25%")))

df_b <- df2 %>% filter(type=="Tumor" & seqnames=="chr1" & compartment=="B") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(type=="Tumor" & seqnames=="chr1" & compartment=="I") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name)) 
coeff$name <- factor(coeff$name, levels=unique(coeff$name))

coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("PC1 Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/eigen_by_hypomethylation_relative_to_minimally_hypometh_tumor.png", width=7, height=4)
```



# HCT116 AZA/DMSO: Prepare a long format eigen/meth data frame
```{r }
comp <- readRDS("../rds/granges_compI_HCT116.rds")
meth_gr <- readRDS("../rds/compartmentGR.rds")

samples <- c('HCT116.5AZA.24h_rep1','HCT116.5AZA.24h_rep2','HCT116.DMSO.24h_rep1','HCT116.DMSO.24h_rep2')
x <- colnames(mcols(meth_gr))
meth <- mcols(meth_gr)[,x[grepl("_opensea", x)]]
colnames(meth) <- sub("_opensea", "", colnames(meth))
eigen <- mcols(meth_gr)[,x[grepl("_eigen", x)]]
colnames(eigen) <- sub("_eigen", "", colnames(eigen))

# Rename meth samples 3,4 to 1,2
colnames(meth)[colnames(meth)=="HCT116.5AZA.24h_rep3"] <- "HCT116.5AZA.24h_rep1"
colnames(meth)[colnames(meth)=="HCT116.5AZA.24h_rep4"] <- "HCT116.5AZA.24h_rep2"
colnames(meth)[colnames(meth)=="HCT116.DMSO.24h_rep3"] <- "HCT116.DMSO.24h_rep1"
colnames(meth)[colnames(meth)=="HCT116.DMSO.24h_rep4"] <- "HCT116.DMSO.24h_rep2"


rds <- paste0("../rds/eigen_hct116_aza_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {
  registerDoParallel(cores=4)
  df <- foreach(sample = samples, .combine=rbind) %dopar% {
    df2 <- foreach (chr=chrs, .combine=rbind) %do% {
      idx <- seqnames(granges(meth_gr))==chr
      gr <- granges(meth_gr)[idx]
      
      if (sample %in% colnames(eigen)) {
        gr$eigen <- eigen[idx, sample]  
      } else {gr$eigen <- NA}
      if (sample %in% colnames(meth)) {
        gr$meth <- meth[idx, sample]  
      } else {gr$meth <- NA}
      
      idx <- findOverlaps(gr, comp, select = "first")
      gr$compartment <- comp$subComp_HCT116[idx]
      
      rep <- str_split_fixed(sample, "_rep", 2)[2]
      
      gr %>% as.data.frame() %>% select(seqnames, start, compartment, eigen, meth) %>% 
        cbind(sample=sample, rep=rep, type=ifelse(grepl("AZA", sample), "AZA", "DMSO"))
    }
  }  
  df$type <- relevel(df$type, ref="DMSO")
  df$compartment <- factor(df$compartment, levels=c("A", "I", "B"))
  saveRDS(df, file=rds)
}
```



# HCT116 Aza PC1 difference (O/E) relative to minimally hypomethylated regions

eigen = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(chr_pos = paste(seqnames, start)) %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[type=="DMSO"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25,  1), labels=c("<15%", "15%-20%", "20%-25%", ">25%")))

df_b <- df2 %>% filter(compartment=="B") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")


coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name))
coeff$name <- factor(coeff$name, levels=unique(coeff$name))
 
coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("PC1 Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/eigen_by_hypomethylation_relative_to_minimally_hypometh_hct116.png", width=7, height=4)
```




# Aging WI-38 fibroblasts: Prepare a long format eigen/meth data frame
```{r }
samples <- c('Wi38_16_1','Wi38_16_2','Wi38_30_1','Wi38_30_2','Wi38_40_1','Wi38_40_2','Wi38_46_1','Wi38_46_2')

rds <- paste0("../rds/eigen_wi38_", bin_size, ".rds")
if (file.exists(rds)) {
  df <- readRDS(rds)
} else {

  comp <- readRDS("../rds/compartmentGR.rds")
  
  # Get eigenvectors (PC1)
  x <- colnames(mcols(comp))
  eigen <- mcols(comp)[,x[grepl("Wi38_.._rep", x)]]
  colnames(eigen) <- sub("_eigen", "", sub("rep", "", colnames(eigen)))

  # Get methylation
  wgbs <- readRDS("../rds/wgbs_iw48.rds")
  time <- sub("p", "", str_split_fixed(sampleNames(wgbs), "_", 4)[,3])
  rep <- str_split_fixed(sampleNames(wgbs), "_", 4)[,4]
  sampleNames(wgbs) <- paste0("Wi38_", time, "_", rep)
  tile_meth <- getMeth(wgbs, regions = comp, type = "raw", what = "perRegion")
  
  # Define compartments based on P16 eigenvector and P40-P16 hypomethylation
  comp$wi38_hypometh <- rowMeans(tile_meth[,c("Wi38_16_1", "Wi38_16_2")]) - tile_meth[,c("Wi38_40_2")]
  comp$wi38_comp <- ifelse(comp$Wi38_16_mean_eigen>0, "A", "B")
  comp$wi38_comp[comp$Wi38_16_mean_eigen>0 & comp$wi38_hypometh>0.1] <- "I"
  #table(comp$wi38_comp)
  
  df <- foreach(sample = samples, .combine=rbind) %do% {
      comp %>% as.data.frame() %>% select(seqnames, start) %>% 
        cbind(compartment=comp$wi38_comp, eigen=eigen[,sample], meth=tile_meth[,sample], sample=sample,
              time=str_split_fixed(sample, "_", 3)[,2], rep=str_split_fixed(sample, "_", 3)[,3])
  }  
  saveRDS(df, file=rds)
}

```


# WI-38 fibroblasts P40 vs P16: PC1 relative to minimally hypomethylated regions

eigen = hypometh_bin + b_pos + b_sample 

```{r}
df2 <- df %>% mutate(chr_pos = paste(seqnames, start)) %>%
  filter(time %in% c("16", "40") & sample!="Wi38_40_1") %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation = mean(meth[time=="16"], na.rm=TRUE) - meth) %>% ungroup() %>%
        mutate(hypomethylation_bin=cut(hypomethylation, c(-1, 0.15, 0.2, 0.25, 1), labels=c("<15%", "15%-20%", "20%-25%", ">25%"))) %>%
  ungroup()

df_b <- df2 %>% filter(compartment=="B") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(eigen ~  hypomethylation_bin + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("hypomethylation_bin", name))%>% mutate(name = sub("hypomethylation_bin", "", name))
coeff$name <- factor(coeff$name, levels=unique(coeff$name))
 
coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("Hypomethylation") + ylab("PC1 Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18), axis.text.x = element_text(angle = 90, hjust=1)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/eigen_by_hypomethylation_relative_to_minimally_hypometh_wi38_p40_vs_p16.png", width=7, height=4)
```



# Aging WI-38 fibroblasts by time: PC1 relative to early passage

```{r}
df2 <- df %>% mutate(chr_pos = paste(seqnames, start)) %>%
  filter(time %in% c("16", "30", "40") & sample!="Wi38_40_1") %>%
  group_by(seqnames, start) %>% 
        mutate(hypomethylation40 = mean(meth[time=="16"], na.rm=TRUE) - mean(meth[time=="40"], na.rm=TRUE)) %>% ungroup() %>%
  filter(hypomethylation40 > 0.20)

df_b <- df2 %>% filter(compartment=="B") 
fit <- lmer(eigen ~ time + (1|sample) + (1|chr_pos), data=df_b)
coeff_b <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="B")

df_i <- df2 %>% filter(compartment=="I") 
fit <- lmer(eigen ~  time + (1|sample) + (1|chr_pos), data=df_i)
coeff_i <- fixef(fit) %>% enframe() %>% cbind(confint(fit)[names(fixef(fit)),], compartment="I")

coeff <- rbind(coeff_b, coeff_i)
coeff$compartment <- factor(coeff$compartment, levels=c("I", "B"))
coeff <- coeff %>% filter(grepl("time", name))%>% mutate(name = sub("time", "P", name))

coeff %>% ggplot(aes(name, value, color=compartment)) + geom_point(size=4) + geom_errorbar(aes(ymin=`2.5 %`, ymax=`97.5 %`, color=compartment), size=2) + scaleCompCol + theme_bw()+ xlab("WI-38 Passage Number") + ylab("PC1 Difference") + geom_hline(yintercept = 0, linetype=2) + facet_wrap(~compartment) + theme(legend.position = "none") + theme(text = element_text(size=18)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../plots/eigen_by_time_relative_to_p16_wi38_p40_hypometh_regions.png", width=7, height=4)

```



