---
title: "Enhancer promoter loops in colon tissue and colon normals"
output: html_document
abstract: "This vignette provides the code and processed data objects to reproduce the enhancer-promoter loop analysis of the paper by Johnstone, Reyes. et al 2020."
---


We first load the necessary libraries to complete this vignette. 

```{r, message=FALSE}

library(diffloop)
library(magrittr)
library(genefilter)
library(DESeq2)
library(limma)
library(fdrtool)
library(edgeR)
library(cowplot)
library(gwascat)
library(liftOver)
```

The code below loads pre-computed R objects containing the *HiChIP* counts for each sample as well as the promoter regions of each transcript in the human genome. 

```{r}

##library(ColonDNATopology)

data(loopsAll_hichipper2)
data(promoters_per_transcript)
loopsAll <- loops

promoters <- unlist(range(promoters))
nms <- names(promoters)
promoters <- padGRanges(promoters, pad=1000)
names( promoters ) <- nms
strand( promoters ) <- "*"
mcols(promoters)$gene <- nms

enhancers <-bedToGRanges( system.file( "extdata/bed/union_K27Ac_peaks.bed",  package="ColonDNATopology") )

enhancers <- padGRanges( enhancers, pad = 1000 )
loopsAll <- mangoCorrection( loopsAll )

totLoopNumber <- sum( loopsAll@rowData$mango.FDR < 0.05 & rowSums(loopsAll@counts >= 3) >= 2 )

```

We detect a total of `r format(totLoopNumber, big.mark = ",")` significant loops with enough counts. How many of them are enhancer promoter loops?

```{r}
loopsEP <- ColonDNATopology:::keepEPnew( loopsAll, enhancers, promoters )
cc <- loopsEP@counts
colnames(cc) <- gsub(".filt.intra", "", colnames(cc))
keep <- loopsEP@rowData$mango.FDR<0.05 & rowSums(cc >= 3) >= 2
epLoopNum <- sum(keep)
```

The total number of enhancer-promoter loops is `r format(epLoopNum, big.mark = ",")`. We can now format the loop count matrices into **DESeqDataSet** objects for normalization of both copy numbers, library size and non-linear trends.

```{r}
loopDsd <- DESeqDataSetFromMatrix(
    cc[keep,],
    colData=DataFrame(condition=factor(ifelse( grepl( "N$", colnames( cc ) ), "Normal", "Tumor" ))),
    design=~condition )

data(cnvs_40K)

coords <- strsplit( gsub(":-", ":", as.character( summary(loopsEP)$region[keep] )), ":|-" )
rowRanges(loopDsd) <- GRanges(
    sapply( coords, "[[", 1),
    IRanges(
        start=pmax(0, as.numeric(sapply( coords, "[[", 2 ))),
        end=as.numeric(sapply( coords, "[[", 3)) ) )
keep[which(keep)][as.vector(seqnames( rowRanges(loopDsd) ) %in% "chrX")] <- FALSE
loopDsd <- loopDsd[!seqnames( rowRanges(loopDsd) ) %in% "chrX",]
rownames(loopDsd) <- sprintf("loop%0.5d", seq_len(length(rowRanges(loopDsd))))
loopDsd <- copyNumberNormalize( loopDsd, cnvs )
loopDsd <- normalizeTrend( loopDsd, prevNorm=TRUE )
##loopDsd <- estimateSizeFactors( loopDsd )
##loopDsd <- normalizeTrend( loopDsd, prevNorm=FALSE )
##loopDsd <- DESeq( loopDsd )
groups <- colData( loopDsd )$condition

z <- DGEList(counts = counts( loopDsd ), group = groups)
design <- model.matrix(~groups)
z$offset <- normalizationFactors( loopDsd )
yy <- estimateGLMCommonDisp(z, design)
fit <- glmQLFit(yy, design, robust = TRUE)
qlf <- glmLRT(fit, coef = 2)
results <- as.data.frame(topTags(qlf, n = nrow(z@.Data[[1]]),
                                 sort.by = "none"))
loops <- subsetLoops( loopsEP, keep )
loops <- quickAssoc( loops )

loopSummary <- summary( loops )

data(summaryExprTCGA)
data(summaryExprOurs)

loopSummary$baseMean <- rowMeans( counts( loopDsd, normalized=TRUE ) )
loopSummary$gene.tss <- strsplit( loopSummary$gene.tss, "," )
```

We detect `r sum( loopSummary$FDR < 0.5 & loopSummary$logFC > 0 )` loops that are stronger in tumors compared to normals, and `r sum( loopSummary$FDR < 0.5 & loopSummary$logFC < 0 )` that are wearker in tumors compared to normals colons.

We next stratify loops according to both their mean strength and magnitud of change in tumors compared to normals.

```{r}
loopSummary <- loopSummary %>%
    dplyr::mutate(
               loopCat =
                   dplyr::case_when(
                              FDR < 0.5 & logFC > 0 & baseMean >= 10 ~ "Gained (strong)",
                              FDR < 0.5 & logFC > 0 & baseMean < 10 ~ "Gained (weak)",
                              FDR < 0.5 & logFC < 0 & baseMean >= 10 ~ "Lost (strong)",
                              FDR < 0.5 & logFC < 0 & baseMean < 10 ~ "Lost (weak)",
                              TRUE ~ "No change"
                          ),
               loopCat2 =
                   dplyr::case_when(
                              FDR < 0.5 & logFC > 0 ~ "Gained",
                              FDR < 0.5 & logFC < 0 ~ "Lost",
                              TRUE ~ "No change"
                          )
           )

loopSummary$loopCat <- factor( loopSummary$loopCat,
                              levels=rev(c("Gained (strong)", "Gained (weak)",
                                       "No change", "Lost (weak)",
                                       "Lost (strong)" ) ))
levels( loopSummary$loopCat ) <- c("--", "-", "0", "+", "++")

loops@rowData$loopID <- rownames(loopDsd)
loopSummary$loopID <- rownames(loopDsd)

loopSummary %>%
    dplyr::filter( baseMean > 0, loopCat != "unchanged" ) %>%
    dplyr::group_by( loopCat ) %>%
    dplyr::summarize( dplyr::n() )
kp <- rowSums(loopSummary[,grepl("N.filt.intra", colnames( loopSummary ))] > 1) > 0

meanCounts <- sapply(
    split( seq_len( ncol(loopDsd) ), ifelse( grepl( "N$", colnames( loopDsd ) ), "Normal", "Tumor" ) ),
    function(x){
        rowMeans( (counts(loopDsd, normalized=TRUE) + 1.5)[,x])
    } )
loopSummary$logFC3 <- log2(meanCounts[,"Tumor"]/meanCounts[,"Normal"])

minus_log10 <- scales::trans_new("minus_log10",
       transform=function(x){ -log10(x) },
       inverse=function(x){ 10^{x*-1} } )

library(scales)
library(ggplot2)
theme_set(theme_cowplot())
```

We also plot a volcano plot of the differential analysis of loop strength, which corresponds to Figure 1A. 
 
```{r, fig.height=3.5, fig.width=3.5, out.width = '75%'}
loopSummary %>%
    dplyr::filter( baseMean > 2 ) %>%
    dplyr::select( logFC3, PValue, FDR ) %>%
    ggplot( aes( logFC3, PValue,
                col=dplyr::case_when(
                               FDR < 0.5 & logFC3 > 0 ~ "gained",
                               FDR < 0.5 & logFC3 < 0 ~ "lost",
                               TRUE ~ "non" ) ) ) +
    geom_point(size=.2) + theme(legend.pos="none") +
    scale_color_manual(
        values = c(lost="red", gained="#32CD32", non="black") ) +
    scale_y_continuous(trans=minus_log10,
                       labels=trans_format('log10', math_format(10^.x)),
                       breaks = c(0, 0.1, 0.01, 0.001, 0.0001)) +
    labs(
        y=expression(p*"-value"),
        x=expression("Loop"~log[2]~"(T/N)")
        )
```

Are there gene expression changes associated with the differential loops? The code below shows the gene expression changes of the genes for each of the loop categories defined above (based on average loop strength and magnitud of change in tumors). 

```{r, fig.height=3, fig.width=4.2}
loopSummary %>%
    dplyr::select( gene.tss, logFC, loopCat ) %>%
    tidyr::unnest(cols=c(gene.tss) ) %>%
    dplyr::rename( geneID=gene.tss, loopDiff=logFC ) %>%
    dplyr::full_join( exprTCGA, by="geneID" ) %>%
    dplyr::select( loopDiff, geneID, cancer.means, normal.means, loopCat ) %>%
    dplyr::mutate( log2FC=log2( cancer.means / normal.means ) )  %>%
    na.omit() %>%
    ggplot( aes( loopCat, log2FC, fill=loopCat ) ) +
    geom_boxplot(outlier.shape=NA) +
    geom_hline(yintercept=0, col="darkred") +
    scale_fill_manual(
        values = c(`--`="red", `++`="#32CD32", `+`="#32CD3250", `-`="#ff000050", `0`="white") ) +
    labs(x=expression("Loop"~log[2]~"(T/N)"),
         y=expression("Expression"~log[2]~"(T/N)")) +
    coord_cartesian(ylim=c(-3,3)) +
    theme(legend.pos="none")
```

The plot above shows that as the loops are stronger in tumors, genes tend to be over-expressed compared to normals, and viceversa. The code above uses TCGA gene expression. Is this behaviour recapitulated in the expression data of our cohort?

```{r, fig.height=3, fig.width=4.2}
loopSummary %>%
    dplyr::select( gene.tss, logFC, loopCat ) %>%
    tidyr::unnest(cols=c(gene.tss) ) %>%
    dplyr::rename( geneID=gene.tss, loopDiff=logFC ) %>%
    dplyr::full_join( exprOurs, by="geneID" ) %>%
    dplyr::select( loopDiff, geneID, Tumor, Normal, loopCat ) %>%
    dplyr::mutate( log2FC=log2( Tumor / Normal ) )  %>%
    na.omit() %>%
    ggplot( aes( loopCat, log2FC ) ) +
    geom_boxplot(outlier.shape=NA) +
    geom_hline(yintercept=0, col="darkred") +
    labs(x=expression("Loop"~log[2]~"(T/N)"),
         y=expression("Expression"~log[2]~"(T/N)")) +
    coord_cartesian(ylim=c(-3,3))
```


The global trends are consistent. 

## Table S3

The code below produces the table S3, which also incorporates some of the CNV data. 

```{r}
coords <- strsplit( gsub(":-", ":", as.character( loopSummary$region )), ":|-" )

summaryCoords <- GRanges(
    sapply( coords, "[[", 1),
    IRanges(
        start=pmax(0, as.numeric(sapply( coords, "[[", 2 ))),
        end=as.numeric(sapply( coords, "[[", 3)) ) )

names( summaryCoords ) <- sprintf("loop%0.5d", seq_len(length(summaryCoords)))
ovl <- findOverlaps( summaryCoords, cnvs )
spOvl <- split( subjectHits(ovl), names(summaryCoords)[queryHits(ovl)] )

##x <- spOvl[[1]]

cnvsMat <- as.matrix(mcols(cnvs))

cnvPerLoop <- vapply(spOvl, function( x ){
    colMeans( cnvsMat[x,,drop=FALSE], na.rm=TRUE )
}, numeric( ncol( cnvsMat ) ) )

cnvPerLoop <- as.data.frame(t(cnvPerLoop))
cnvPerLoop <- cnvPerLoop[,colnames(cnvPerLoop) %in% rownames(colData(loopDsd))]

colnames(cnvPerLoop) <- paste0("cnv_", colnames(cnvPerLoop))

cnvPerLoop$loopID <- rownames( cnvPerLoop )
loopSummary$loopID <- rownames( cnvPerLoop )

consistentFC <-
    loopSummary %>%
    dplyr::filter( baseMean > 0, loopCat != "unchanged" ) %>%
    dplyr::select( loopID, chr_1, start_1, end_1, chr_2, start_2, end_2,
                  baseMean, logFC, FDR, loopCat2, gene.tss ) %>%
    dplyr::rename( geneID = gene.tss, loopFC=logFC, loopFDR=FDR, loopCat=loopCat2 ) %>%
    tidyr::unnest(cols = c(geneID)) %>%
    dplyr::full_join( exprTCGA, by="geneID" ) %>%
    dplyr::select( loopID, chr_1, start_1, end_1, chr_2, start_2,
                  end_2, baseMean.x, loopFC, loopFDR, loopCat, geneID,
                  cancer.means, normal.means, padj.de, lfc.de ) %>%
    dplyr::rename( baseMean = baseMean.x ) %>%
    dplyr::mutate( log2FC=log2( cancer.means / normal.means ) ) %>%
    dplyr::rename( cancerExpr=cancer.means, normalExpr=normal.means,
                  exprFDR=padj.de, exprFC=lfc.de ) %>%
    dplyr::select( -log2FC ) %>%
    dplyr::filter( loopFDR < 0.5, exprFDR < 0.01 ) %>%
    dplyr::filter( (grepl("Gained", loopCat) & exprFC > 0)| (grepl("Lost", loopCat) & exprFC < 0) ) %>%
    dplyr::arrange( desc( abs( exprFC ) ) )

consistentFC <- dplyr::left_join( consistentFC, cnvPerLoop, by="loopID" )


lpID <- consistentFC[consistentFC$geneID %in% "EPHA2",]
reshape2::melt( counts( loopDsd, normalized=TRUE )[rownames( loopDsd ) %in% lpID,] )

lpID <- consistentFC[consistentFC$geneID %in% "PDCD4",]
reshape2::melt( counts( loopDsd, normalized=TRUE )[rownames( loopDsd ) %in% lpID,] )

##write.table(
##    consistentFC, quote=FALSE, col.names=TRUE,
##    row.names=FALSE, sep="\t",
##    file="epLoops_tumor_normal.txt" )

tabS <- dplyr::select( consistentFC[consistentFC$baseMean > 5,],
                      loopID, chr_1, start_1, end_1, chr_2,
                      start_2, end_2, loopFC, geneID, exprFC )

cnts <- counts( loopDsd, normalized=TRUE )[unique(tabS$loopID),]
cnts <- as.data.frame( cnts )
colnames( cnts ) <- paste0("normCounts_", colnames(cnts))
cnts$loopID <- rownames( cnts )
tabS <- dplyr::full_join( tabS, cnts )
colnames(cnvPerLoop) <- gsub("cnv", "CN", colnames(cnvPerLoop))
tabS <- dplyr::left_join( tabS, cnvPerLoop, by="loopID" )
all( cnvPerLoop$loopID == loopSummary$loopID )
head( tabS )
##write.csv( tabS, quote=FALSE, row.names=FALSE, file="../output/tables/tabS_diffloops.csv" )

```

How would the gene expression plot above loop if we only keep CNV-stable regions?

```{r, fig.height=3, fig.width=4.2}

keep <- rowSums( cnvPerLoop[,grepl( "CN", colnames( cnvPerLoop ) )] > 0.8 & cnvPerLoop[,grepl( "CN", colnames( cnvPerLoop ) )] < 1.1 ) == 7
loopSummary %>%
    dplyr::filter( keep ) %>%
    dplyr::select( gene.tss, logFC, loopCat ) %>%
    tidyr::unnest(cols=c("gene.tss")) %>%
    dplyr::rename( geneID=gene.tss, loopDiff=logFC ) %>%
    dplyr::full_join( exprTCGA, by="geneID" ) %>%
    dplyr::select( loopDiff, geneID, cancer.means, normal.means, loopCat ) %>%
    dplyr::mutate( log2FC=log2( cancer.means / normal.means ) )  %>%
    na.omit() %>%
    ggplot( aes( loopCat, log2FC ) ) +
    geom_boxplot(outlier.shape=NA) +
    geom_hline(yintercept=0, col="darkred") +
    labs(x=expression("Loop"~log[2]~"(T/N)"),
         y=expression("RNA"~log[2]~"(T/N)")) +
    coord_cartesian(ylim=c(-3,3))

```


## Overlap of enhancers with SNPs

```{r}

###########################################
## Are promoters enriched among CRC SNPs ##
###########################################

data("crcSNPs")
data("promoters_per_transcript")

promoters <- reduce( unlist( promoters ), ignore.strand=TRUE )
promoters <- keepStandardChromosomes( promoters, pruning.mode="coarse" )

enhancersWSNP <- subsetByOverlaps( enhancers, crcSNPs )
ovr <- findOverlaps( enhancersWSNP, crcSNPs )
mcols( enhancersWSNP )$rsIDs <- split( names(crcSNPs)[subjectHits( ovr )], queryHits( ovr ) )

loopsSub <- loops
sumLoops <- summary( loopsSub )

left <- unique(subsetByOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), crcSNPs ))
right <- unique(subsetByOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), crcSNPs ))
candAnchors <- unique(c( left, right ))

ovr <- findOverlaps( candAnchors, crcSNPs )
mcols( candAnchors )$rsIDs <- split( names(crcSNPs)[subjectHits( ovr )], queryHits( ovr ) )

library(ggpubr)
library(ensembldb)
library(ggbio)
library(Rsamtools)
library(GenomicAlignments)

data("tcgaExpressionCR_hg19")
eDat <- rowRanges( tcgaExpressionCR_hg19 )

geneRange <- unlist( range( eDat ) )
```

The code below plots a loop that conects an enhancer element to the promoter of the *COLCA1* and *COLCA2* genes. The enhancer overlaps with a SNP that is a hit of predisposition to colorectal cancer. 

```{r, fig.height=4.5, fig.width=5}
### COLCA Loci
i <- 10
##chr11 111168244-111172353      * | rs3802842
enh <- candAnchors[i]
ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), enh )
ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), enh )
ovr <- c( queryHits(ovr1), queryHits(ovr2) )
x <- sumLoops[ovr,]
spReg <- strsplit( as.character( x$region ), ":|-" )
spReg <- GRanges( sapply( spReg, "[[", 1 ),
                 IRanges( as.numeric( sapply( spReg, "[[", 2 ) ), as.numeric( sapply( spReg, "[[", 3 ) ) ) )
spReg <- range( spReg )
gns <- unique(unlist(strsplit( x$gene.tss, "," )))
plotGR <-
    range( c( resize( enh, width=width(enh)+1000, fix="center" ),
             resize( geneRange[gns], width( geneRange[gns] ) + 2000, fix="center" ),
             spReg
             ),
          ignore.strand=TRUE )
start(plotGR) <- 111152500
end( plotGR ) <- 111180000
rmtheme <- theme(legend.pos="none",
                 axis.text.x=element_blank(),
                 axis.title.x=element_blank(),
                 plot.margin = margin(0, 0, .1, 0, "cm"))
chr <- unique( as.character(x$chr_1) )
lop <-
    GRanges(chr,     
            IRanges(
                start(resize( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), 1, fix="center" )),
                start( resize( GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ), 1, fix="center" ) ) ) )
lop <- subsetByOverlaps( lop, plotGR )
enhInAnch <- subsetByOverlaps( enhancers, enh )
plotGR <- keepStandardChromosomes( plotGR )

plot_gr <- plotGR

ac <- plotCovProfileForSample( 
  sampleName=c("BRD3162","BRD3179N"),
  protein="K27Ac", plotGR,
  sampleGroups=c("Tumor", "Normal"), itersmooth=1,
  meanSmooth=TRUE, lwd=1 ) +
  rmtheme +
  scale_color_manual( values=c( `Tumor` = "#984ea399", `Normal` = "#4daf4a99") )

ap <- plotLoopsFromRanges( lop, plot_gr=plotGR ) +
    geom_rect( data = as.data.frame( enhInAnch ),
              inherit.aes=FALSE,
              aes( xmin=start, xmax=end, ymin=-Inf, ymax=Inf ),
              fill="#cc4c02", alpha=0.35 ) +
    geom_vline( xintercept=start( subsetByOverlaps(unlist(crcSNPs), enh) ), colour="#377eb8", size=0.5) +
    theme( axis.line.y=element_blank(), axis.ticks.y=element_blank() )

bp <- plotGenesFromGRangesList( eDat, plotGR )
abp <- ggarrange( ap + rmtheme,
                 ac,
                 bp + theme(legend.pos="none") +
                 scale_x_continuous(labels=function(x){round(x/1000000, 2)}) +
                 xlab("Genome (Mb)"), nrow=3, align="v", heights=c( 0.3, 0.4, 1 ) )
abp
```

The code below plots the equivalent plot but for a different locus. These data shows an enhancer of the gene TERC that overlaps with a colorectal cancer GWAS SNP.

```{r, fig.height=4.5, fig.width=5}

i <- 19
## chr3 169486553-169493266      * | rs10936599
enh <- candAnchors[i]

ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), enh )
ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), enh )
ovr <- c( queryHits(ovr1), queryHits(ovr2) )
x <- sumLoops[ovr,]
spReg <- strsplit( as.character( x$region ), ":|-" )
spReg <- GRanges( sapply( spReg, "[[", 1 ),
                 IRanges( as.numeric( sapply( spReg, "[[", 2 ) ), as.numeric( sapply( spReg, "[[", 3 ) ) ) )
spReg <- range( spReg )
gns <- unique(unlist(strsplit( x$gene.tss, "," )))
plotGR <-
    range( c( resize( enh, width=width(enh)+1000, fix="center" ),
             resize( geneRange[gns], width( geneRange[gns] ) + 2000, fix="center" ),
             spReg
             ),
          ignore.strand=TRUE )
start(plotGR) <- 169480500
end( plotGR ) <- 169508000
rmtheme <- theme(legend.pos="none",
                 axis.text.x=element_blank(),
                 axis.title.x=element_blank(),
                 plot.margin = margin(0, 0, .1, 0, "cm"))
chr <- unique( as.character(x$chr_1) )
lop <-
    GRanges(chr,     
            IRanges(
                start(resize( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), 1, fix="center" )),
                start( resize( GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ), 1, fix="center" ) ) ) )
lop <- subsetByOverlaps( lop, plotGR )
enhInAnch <- subsetByOverlaps( enhancers, enh )
enhInAnch <- resize( enhInAnch, round(width(enhInAnch) * 1.2), fix="center")
if( length( enhInAnch ) == 0 ){ next }
plotGR <- keepStandardChromosomes( plotGR )
### plot loops from the middle of the K27Ac peaks instead of middle of the anchor ##
enhancersHere <- resize( subsetByOverlaps( enhancers, plotGR ), 1, fix="center" )
lop <- GRanges(chr, IRanges( start( enhancersHere[1] ), end( enhancersHere[3]) ) )

ac <- plotCovProfileForSample( sampleName=c("BRD3162","BRD3179N"),
                              protein="K27Ac", plotGR,
                              sampleGroups=c("Tumor", "Normal"), itersmooth=1,
                              meanSmooth=TRUE, lwd=1 ) +
    rmtheme +
    scale_color_manual( values=c( `Tumor` = "#984ea399", `Normal` = "#4daf4a99") )

ap <- plotLoopsFromRanges( lop, plot_gr=plotGR ) +
     ## geom_rect( data = as.data.frame( enh ), inherit.aes=FALSE,
     ##           aes( xmin=start, xmax=end, ymin=-Inf, ymax=Inf ),
    ##           fill="gray", alpha=0.35 ) +
    geom_rect( data = as.data.frame( enhInAnch[2] ),
              inherit.aes=FALSE,
              aes( xmin=start, xmax=end, ymin=-Inf, ymax=Inf ),
              fill="#cc4c02", alpha=0.35 ) +
    geom_vline( xintercept=start( subsetByOverlaps(unlist(crcSNPs), enh) ), colour="#377eb8", size=0.5) +
    theme( axis.line.y=element_blank(), axis.ticks.y=element_blank() )
bp <- plotGenesFromGRangesList( eDat, plotGR, gpSelf=FALSE )
abp <- ggarrange( ap + rmtheme,
                 ac,
                 bp + theme(legend.pos="none") +
                 scale_x_continuous(labels=function(x){round(x/1000000, 3)}) +
                 xlab("Genome (Mb)"), nrow=3, align="v", heights=c( 0.3, 0.4, 1 ) )
abp

```

Finally, the code below repeats this plot but for the CXCR4 locus. 

```{r}

i <- 18
## chr3 169486553-169493266      * | rs10936599
enh <- candAnchors[i]
ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), enh )
ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), enh )
ovr <- c( queryHits(ovr1), queryHits(ovr2) )
x <- sumLoops[ovr,]
spReg <- strsplit( as.character( x$region ), ":|-" )
spReg <- GRanges( sapply( spReg, "[[", 1 ),
                 IRanges( as.numeric( sapply( spReg, "[[", 2 ) ), as.numeric( sapply( spReg, "[[", 3 ) ) ) )
spReg <- range( spReg )
gns <- unique(unlist(strsplit( x$gene.tss, "," )))
plotGR <-
    range( c( resize( enh, width=width(enh)+1000, fix="center" ),
             resize( geneRange[gns], width( geneRange[gns] ) + 2000, fix="center" ),
             spReg
             ),
          ignore.strand=TRUE )
start(plotGR) <- 136860000
end( plotGR ) <- 137005000
rmtheme <- theme(legend.pos="none",
                 axis.text.x=element_blank(),
                 axis.title.x=element_blank(),
                 plot.margin = margin(0, 0, .1, 0, "cm"))
chr <- unique( as.character(x$chr_1) )
lop <-
    GRanges(chr,     
            IRanges(
                start(resize( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), 1, fix="center" )),
                start( resize( GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ), 1, fix="center" ) ) ) )
lop <- subsetByOverlaps( lop, plotGR )
enhInAnch <- subsetByOverlaps( enhancers, enh )
enhInAnch <- resize( enhInAnch, round(width(enhInAnch) * 1.2), fix="center")
if( length( enhInAnch ) == 0 ){ next }
plotGR <- keepStandardChromosomes( plotGR )
ac <- plotCovProfileForSample( sampleName=c("BRD3162" ),
                              protein="K27Ac", plotGR,
                              sampleGroups=c("Tumor"), itersmooth=1,
                              meanSmooth=TRUE, lwd=1 ) +
    rmtheme +
    scale_color_manual( values=c( `Tumor` = "#984ea399", `Normal` = "#4daf4a99") )
ap <- plotLoopsFromRanges( lop, plot_gr=plotGR ) +
      ## geom_rect( data = as.data.frame( enh ), inherit.aes=FALSE,
      ##           aes( xmin=start, xmax=end, ymin=-Inf, ymax=Inf ),
      ##          fill="#cc4c02", alpha=0.35 ) +
    geom_rect( data = as.data.frame( enhInAnch ),
              inherit.aes=FALSE,
              aes( xmin=start, xmax=end, ymin=-Inf, ymax=Inf ),
              fill="#cc4c02", alpha=0.35 ) +
    geom_vline( xintercept=start( subsetByOverlaps(unlist(crcSNPs), enh) ), colour="#377eb8", size=0.5) +
    theme( axis.line.y=element_blank(), axis.ticks.y=element_blank() )
bp <- plotGenesFromGRangesList( eDat, plotGR, gpSelf=FALSE )
abp <- ggarrange( ap + rmtheme,
                 ac,
                 bp + theme(legend.pos="none") +
                 scale_x_continuous(labels=function(x){round(x/1000000, 3)}) +
                 xlab("Genome (Mb)"), nrow=3, align="v", heights=c( 0.3, 0.4, 1 ) )
abp
```

## Differential loops (examples)

The manuscript highlights plots of individual E-P loops that are gained or lost, which are associated to differences in gene expression patterns. The code below reproduces the plot from Figure 1, where we display a differential loop that is strongly associated to the expression of the gene *EPHA2*. The plot below shows the loop counts.

```{r, fig.width=3.4, fig.height=3.4}

plotGR <- range( eDat[["EPHA2"]] )
ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), plotGR )
ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), plotGR )
ovr <- unique(c( queryHits(ovr1), queryHits(ovr2) ))
x <- sumLoops[ovr,]
spReg <- strsplit( as.character( x$region ), ":|-" )
spReg <- GRanges( sapply( spReg, "[[", 1 ),
                 IRanges( as.numeric( sapply( spReg, "[[", 2 ) ), as.numeric( sapply( spReg, "[[", 3 ) ) ) )
spReg <- range( spReg )
gns <- unique(unlist(strsplit( x$gene.tss, "," )))

plotGR <-
    range( c( plotGR,
             resize( geneRange[gns], width( geneRange[gns] ) + 2000, fix="center" ),
             spReg
             ),
          ignore.strand=TRUE )
start(plotGR) <- 16405000
end(plotGR) <-   16522000

rmtheme <- theme(legend.pos="none",
                 axis.text.x=element_blank(),
                 axis.title.x=element_blank(),
                 plot.margin = margin(0, 0, .1, 0, "cm"))
chr <- unique( as.character(x$chr_1) )
keep <- rowMeans( counts( loopDsd, normalized=TRUE )[x$loopID,] ) > 5
x <- x[keep,]

##png("../output/plots/jitter_loopCounts_EPHA2.png", width=2.4, height=2.4, unit="in", res=300)
as.data.frame( counts(loopDsd, normalized=TRUE)[rownames(loopDsd) %in% x$loopID,] ) %>%
    tibble::rownames_to_column(var="loopID") %>%
    reshape2::melt() %>%
    dplyr::filter( loopID %in% "loop00221" ) %>%
    dplyr::mutate( sampType=ifelse( grepl("N$", variable ), "Normal", "Tumor" ) ) %>%
    ggplot( aes( sampType, log2(value+1), col=sampType ) ) +
    geom_point(alpha=0.8, size=0.8) +
    scale_color_manual(values=c(`Normal`="#4daf4a", `Tumor`="#984ea3")) +
    theme( legend.pos="none" ) +
    labs( y="EPHA2 loop\n(normalized counts)", x="" )
##dev.off()

```

## *EPHA2* differential loop

The code below shows the E-P loops that link enhancers to the promoter of the *EPHA2* gene.  

```{r, fig.height=3.8, fig.width=4.2}

anchors <- c( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ) )
lop <-
    GRanges(chr,     
            IRanges(
                start(resize( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), 1, fix="center" )),
                start( resize( GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ), 1, fix="center" ) ) ) )
lop <- subsetByOverlaps( lop, plotGR )

data(k27ac_data)

#k27Peaks <- readRDS("/data/aryee/bernstein/chip/SummarizedData/gcapc/peaks/mergeLibPeaks_K27Ac.rds")
k27Peaks <- k27Peaks[seqnames( k27Peaks ) != "chrY",]
names( k27Peaks ) <- rownames( k27Counts )

ovl1 <- findOverlaps( k27Peaks, GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ) )
ovl1 <- split( k27Counts[queryHits( ovl1 ),"BRD3162-K27Ac"], subjectHits( ovl1 ) )
ovl2 <- findOverlaps( k27Peaks, GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ) )
ovl2 <- split( k27Counts[queryHits( ovl2 ),"BRD3162-K27Ac"], subjectHits( ovl2 ) )
lop <- GRanges(
    x$chr_1,
    IRanges( 
        start( resize( k27Peaks[vapply( ovl1, function(x){ names(x)[which.max(x)] }, character(1)),], 1, fix="center" ) ),
        start( resize( k27Peaks[vapply( ovl2, function(x){ names(x)[which.max(x)] }, character(1)),], 1, fix="center" ) ) ) )

### chose highest peak ###

plotGR <- keepStandardChromosomes( plotGR )
ac <- plotCovProfileForSample( sampleName=c("BRD3162", "BRD3179N"),
                              protein="K27Ac", plotGR,
                              sampleGroups=c("Tumor", "Normal"), itersmooth=1,
                              meanSmooth=TRUE, lwd=1 ) +
    rmtheme +
    scale_color_manual( values=c( `Tumor` = "#984ea399", `Normal` = "#4daf4a99") )

valuesTumors <- rowMedians( counts(loopDsd, normalized=TRUE)[x$loopID,colData( loopDsd )$condition == "Tumor",drop=FALSE] )
valuesNormal <- rowMedians( counts(loopDsd, normalized=TRUE)[x$loopID,colData( loopDsd )$condition == "Normal",drop=FALSE] )
maxVal <- max(c(valuesTumors, valuesNormal))
minVal <- min(c(valuesTumors, valuesNormal))
aTumor <- plotLoopsFromRanges( lop, plot_gr=plotGR, rangeSize=c(0, 1),
                              valueVars=valuesTumors, widthLims=c(2, maxVal), sd=40 )
aNormal <- plotLoopsFromRanges( lop, plot_gr=plotGR, rangeSize=c(0, 1),
                               valueVars=valuesNormal, widthLims=c(2, maxVal), sd=40 )
bp <- plotGenesFromGRangesList( eDat, plotGR, gpSelf=FALSE )
abp <- ggarrange(
    aNormal +
    rmtheme +
    ylab("N") +
    theme( axis.ticks.y=element_blank(), axis.line.y=element_blank(),
          axis.title.y=element_text(angle=0, colour="#4daf4a") ),
    aTumor +
    rmtheme + 
    ylab("T") +
    theme( axis.ticks.y=element_blank(), axis.line.y=element_blank(),
          axis.title.y=element_text(angle=0, colour="#984ea3") ),
    ac,
    bp + theme(legend.pos="none") +
    scale_x_continuous(labels=function(x){round(x/1000000, 3)}) +
    xlab("Genome (Mb)"), nrow=4, align="v", heights=c( 0.3, 0.3, 0.4, 0.6 ) )
abp


```

## *PDCD4* differential loop

The plot below shows E-P loops that link enhancers to the promoter of the gene *PDCD4*. 

```{r, fig.width=3.4, fig.height=3.4}
plotGR <- range( eDat[["PDCD4"]] )
ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), plotGR )
ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), plotGR )
ovr <- unique(c( queryHits(ovr1), queryHits(ovr2) ))
x <- sumLoops[ovr,]

spReg <- strsplit( as.character( x$region ), ":|-" )
spReg <- GRanges( sapply( spReg, "[[", 1 ),
                 IRanges( as.numeric( sapply( spReg, "[[", 2 ) ), as.numeric( sapply( spReg, "[[", 3 ) ) ) )
spReg <- range( spReg )
gns <- unique(unlist(strsplit( x$gene.tss, "," )))
plotGR <-
    range( c( plotGR,
             resize( geneRange[gns], width( geneRange[gns] ) + 2000, fix="center" ),
             spReg
             ),
          ignore.strand=TRUE )
start(plotGR) <- 112560000
end(plotGR) <- 112658000
rmtheme <- theme(legend.pos="none",
                 axis.text.x=element_blank(),
                 axis.title.x=element_blank(),
                 plot.margin = margin(0, 0, .1, 0, "cm"))
chr <- unique( as.character(x$chr_1) )
keep <- rowMeans( counts( loopDsd, normalized=TRUE )[x$loopID,] ) > 3
x <- x[keep,]

##png("../output/plots/jitter_loopCounts_PDCD4.png", width=2.4, height=2.4, unit="in", res=300)
as.data.frame( counts(loopDsd, normalized=TRUE)[rownames(loopDsd) %in% x$loopID,] ) %>%
    tibble::rownames_to_column(var="loopID") %>%
    reshape2::melt() %>%
    dplyr::filter( loopID %in% "loop06256" ) %>%
    dplyr::mutate( sampType=ifelse( grepl("N$", variable ), "Normal", "Tumor" ) ) %>%
    ggplot( aes( sampType, log2(value), col=sampType ) ) +
    geom_point(alpha=0.8, size=0.8) +
    scale_color_manual(values=c(`Normal`="#4daf4a", `Tumor`="#984ea3")) +
    theme( legend.pos="none" ) +
    labs( y="PDCD4 loop \n(normalized counts)", x="" )
##dev.off()
```

```{r, fig.height=3.8, fig.width=4.2}

anchors <- c( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ),
             GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ) )
lop <-
    GRanges(chr,     
            IRanges(
                start(resize( GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ), 1, fix="center" )),
                start( resize( GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ), 1, fix="center" ) ) ) )

ovl1 <- findOverlaps( k27Peaks, GRanges( x$chr_1, IRanges( x$start_1, x$end_1 ) ) )
ovl1 <- split( k27Counts[queryHits( ovl1 ),"BRD3162-K27Ac"], subjectHits( ovl1 ) )
ovl2 <- findOverlaps( k27Peaks, GRanges( x$chr_2, IRanges( x$start_2, x$end_2 ) ) )
ovl2 <- split( k27Counts[queryHits( ovl2 ),"BRD3162-K27Ac"], subjectHits( ovl2 ) )
lop <- GRanges(
    x$chr_1,
    IRanges( 
        start( resize( k27Peaks[vapply( ovl1, function(x){ names(x)[which.max(x)] }, character(1)),], 1, fix="center" ) ),
        start( resize( k27Peaks[vapply( ovl2, function(x){ names(x)[which.max(x)] }, character(1)),], 1, fix="center" ) ) ) )

lop <- subsetByOverlaps( lop, plotGR )
plotGR <- keepStandardChromosomes( plotGR )
ac <- plotCovProfileForSample( sampleName=c("BRD3162", "BRD3179N"),
                              protein="K27Ac", plotGR,
                              sampleGroups=c("Tumor", "Normal"), itersmooth=1,
                              meanSmooth=TRUE, lwd=1 ) +
    rmtheme +
    scale_color_manual( values=c( `Tumor` = "#984ea399", `Normal` = "#4daf4a99") )
valuesTumors <- rowMedians( counts(loopDsd, normalized=TRUE)[x$loopID,colData( loopDsd )$condition == "Tumor",drop=FALSE] )
valuesNormal <- rowMedians( counts(loopDsd, normalized=TRUE)[x$loopID,colData( loopDsd )$condition == "Normal",drop=FALSE] )
maxVal <- max(c(valuesTumors, valuesNormal))
minVal <- min(c(valuesTumors, valuesNormal))
aTumor <- plotLoopsFromRanges( lop, plot_gr=plotGR, rangeSize=c(0, 1),
                              valueVars=valuesTumors, widthLims=c(2, maxVal), sd=40 )
aNormal <- plotLoopsFromRanges( lop, plot_gr=plotGR, rangeSize=c(0, 1),
                               valueVars=valuesNormal, widthLims=c(2, maxVal), sd=40 )
bp <- plotGenesFromGRangesList( eDat, plotGR, gpSelf=FALSE )
abp <- ggarrange(
    aNormal +
    rmtheme +
    ylab("N") +
    theme( axis.ticks.y=element_blank(), axis.line.y=element_blank(),
          axis.title.y=element_text(angle=0, colour="#4daf4a") ),
    aTumor +
    rmtheme + 
    ylab("T") +
    theme( axis.ticks.y=element_blank(), axis.line.y=element_blank(),
          axis.title.y=element_text(angle=0, colour="#984ea3") ),
    ac,
    bp + theme(legend.pos="none") +
    scale_x_continuous(labels=function(x){round(x/1000000, 3)}) +
    xlab("Genome (Mb)"), nrow=4, align="v", heights=c( 0.3, 0.3, 0.4, 0.6 ) )
abp
```


## Table supplementary S4

```{r}

data(promoters_per_transcript)

promoters
#promoters2 <- readRDS("/data/aryee/areyes/cimp_hichip/data/chipseq/promoters.rds")
promoters <- resize( promoters, 1500, fix="center")

candDF <- lapply( seq_along( candAnchors ), function(x){
    enh <- candAnchors[x]
    ovr1 <- findOverlaps( GRanges( sumLoops$chr_1, IRanges( sumLoops$start_1, sumLoops$end_1 ) ), enh )
    ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), enh )
    sumSub <- sumLoops[queryHits(ovr1),]
    ovr3 <- findOverlaps( GRanges( sumSub$chr_2, IRanges( sumSub$start_2, sumSub$end_2 ) ), promoters )
    distalGenes1 <- names(promoters)[subjectHits(ovr3)]
    loopID1 <- sumSub$loopID[queryHits( ovr3 )]
    ovr2 <- findOverlaps( GRanges( sumLoops$chr_2, IRanges( sumLoops$start_2, sumLoops$end_2 ) ), enh )
    sumSub <- sumLoops[queryHits(ovr2),]
    ovr3 <- findOverlaps( GRanges( sumSub$chr_1, IRanges( sumSub$start_1, sumSub$end_1 ) ), promoters )
    distalGenes2 <- names(promoters)[subjectHits(ovr3)]
    loopID2 <- sumSub$loopID[queryHits(ovr3)]
    distalGenes <- c( distalGenes1, distalGenes2 )
    loopIDs <- c( loopID1, loopID2 )
    numPeaks <- length( subsetByOverlaps( enhancers, enh ) )
    snpInPeak <- length( subsetByOverlaps( subsetByOverlaps( enhancers, enh ), crcSNPs ) )
    data.frame(
        K27Peaks=numPeaks,
        rsID=paste( unlist( enh$rsIDs ), collapse="," ),
        snpInPeak=numPeaks,
        distalGenes=distalGenes,
        loopID=loopIDs)
} )

candDF <- do.call(rbind, candDF)


candDF$rsID <- strsplit( as.character( candDF$rsID ), "," )
candDF <- tidyr::unnest( candDF, cols=c(rsID))

##crcSNPs <- unlist( crcSNPs, use.names=FALSE )

promoters <- unlist( promoters )
mcols(promoters)$geneID <- sapply( strsplit( names(promoters), "\\." ), "[[", 1 )
names(crcSNPs) <- mcols(crcSNPs)$rsID
distToNear <- distanceToNearest( crcSNPs[candDF$rsID,], promoters )

candDF$nearestGene <- mcols(promoters)$geneID[subjectHits( distToNear )]

candDF <- dplyr::left_join( candDF, sumLoops[,c("chr_1", "start_1", "end_1", "chr_2", "start_2", "end_2", "loopID" )], by="loopID")
candDF$K27Peaks <- NULL
candDF$snpInPeak <- NULL

snpsWDistal <- unique(candDF$rsID[!candDF$distalGenes == candDF$nearestGene])
snpsWClose <- unique(candDF$rsID[candDF$distalGenes == candDF$nearestGene])


length( snpsWDistal[snpsWDistal %in% snpsWClose] )/length(unique(candDF$rsID))
length( snpsWDistal[snpsWDistal %in% snpsWClose] )

length(snpsWDistal)/length(unique(candDF$rsID))

colnames( candDF )[2] <- "Putative target gene"
colnames( candDF )[4] <- "Nearest gene"

candDF <- candDF[,c("rsID", "Nearest gene", "Putative target gene", "loopID", "chr_1", "start_1", "end_1", "chr_2", "start_2", "end_2")]

head( candDF )
```

